Estimating duration in the presence of misclassification

### 2024-03-31 ATACCC.tex
add extra gof plots for ATACCC
proof

First, I will explain how biomarkers are typically modelled, both for SARS-CoV-2 and historically in the biostatistical literature (\cref{ATACCC:sec:background}).
Next, I describe the model of \textcite{hakkiOnset} in detail, which applies these methods to the ATACCC data (\cref{ATACCC:sec:hakki}), before modifying its observation model to improve the robustness of the inference (\cref{ATACCC:sec:observation-modification}).
This modification is based on an earlier analysis using a preliminary version of the dataset, described in \cref{ATACCC:sec:original-analysis}.
I convert the model and parameter estimates of the biomarker's trajectory to a duration of RT-PCR positivity (\cref{ATACCC:sec:duration}).
Finally, I discuss the results (\cref{ATACCC:sec:discussion}).
I will refine this estimate in the following chapters before its use in estimating transmission in \cref{E-transmission}.

@@

Random effects represent the within-individual correlation while borrowing information between individuals~\autocite[chapter 24]{lashModern}.
Measurements of viral load from the same individual are correlated.
A random effect means that these measurements share parameter(s), providing correlation.
However, the random effect(s) for each individual is drawn towards a common mean, allowing information on their value to be shared between individuals.

Following much of the literature mentioned previously, I will take a Bayesian approach in this chapter.
Bayesians often refer to random effect models as hierarchical or multilevel models because they can be interpreted as placing a prior at two levels of a parameter hierarchy~\autocite[chapter 5]{gelmanBDA}.
% $\phi_j$ would be referred to as the bottom level of the hierarchy, with $\mu$ and $\sigma$ being the top level.
% More complex models could have more levels, for instance if the individuals are grouped into subpopulations.

@@

Several features of the \textcite{hakkiOnset} model can be motivated by the desire to apply Stan, a modern and convenient MCMC implementation of the NUTS algorithm (see \cref{E-MCMC}).
The first of these is using a continuous viral load trajectory rather than the more common piecewise linear model.
As NUTS uses gradient information, it is more efficient when the model is differentiable.
This leads to the following model for the viral load (see \cref{ATACCC:fig:viral-load-model} for examples):
\begin{align}
v_i(t) = v_{\max,i} \frac{a_i+b_i}{b_ie^{-a_i(t-\tau_{\max,i})} + a_ie^{b_i(t-\tau_{max,i})}}. \label{ATACCC:eq:viral-load}
\end{align}
Time, $t$, is continuous and defined uniquely for each individual such that $t = 0$ at the time that individual $i$ was first tested.
$v_{\max,i}$ is the maximum viral load, which occurs at time $\tau_{\max,i}$.
As $\lvert t - \tau_{\max} \rvert$ becomes large, the viral load trajectory tends towards exponential growth (before the peak) or decline (after the peak) in viral load at rates $a_i$ and $b_i$ respectively.
% Before the peak, this is because $b(t-\tau_{\max,i})$ is small and hence $a\exp(b(t-\tau_{\max,i}))$ is negligible.
% After the peak, this is because $b\exp(-a(t-\tau_{\max,i}))$ is negligible.
This happens quickly for realistic values of the parameters.
Therefore, the log viral load is approximately piecewise linear (see \cref{ATACCC:fig:viral-load-model} for representative viral load trajectories).
$\vec{\phi}_i = (v_{\max,i}, a_i, b_i)^T$ are the random effects for individual $i$.

@@

The noise in the RT-PCR is modelled by assuming each measurement of log viral load has an iid error distributed $\N(0, \sigma_v^2)$.
Measurements are left censored at the limit of detection, $\vlod$, meaning that any observation which would give a log viral load of less than $\vlod$ is instead observed as negative.
False positives and negatives are modelled by allowing a test failure with probability $\rho$.
Test failures are modelled as producing a log viral load reading drawn from $\N(x_0, \sigma_0^2)$.
These readings are censored in the same way as true readings.
A censored test failure represents a false negative, otherwise the test failure represents a false positive.
\textcite{hakkiOnset} uses a prior for $x_0$ centred on $\vlod$, which encodes a prior belief that false negatives and false positives occur with equal probability.

Let the observed log viral load for individual $i$ at time $t$ be $y_{i,t}$, with negative tests signified by $y_{i,t}=v_\text{lod}$.
Then, the likelihood contribution of this observation is:
\begin{align}
p(y_{i,t} \mid \cdot) &= \begin{cases}
    \rho f_N(y_{i,t} \mid x_0, \sigma_0^2) + (1 - \rho) f_N(y_{i,t} \mid \log v_i(t), \sigma_v^2) & \text{if $y_{i,t} > v_\text{lod}$}\\
    \rho F_N(v_\text{lod} \mid x_0, \sigma_0^2) + (1 - \rho) F_N(\vlod \mid \log v_i(t), \sigma_v^2) & \text{if $y_{i,t} = v_\text{lod}$} \label{ATACCC:eq:hakki-likelihood}
\end{cases}
\end{align}
where $f_N(y \mid \mu, \sigma^2)$ and $F_N(y \mid \mu, \sigma^2)$ are the pdf and cdf at $y$ of a normal distribution with mean $\mu$ and variance $\sigma^2$ respectively.
The first term on each line is the likelihood contribution from test failures, either a false positive (top line) or false negative (bottom line).
The second term on the first line is the likelihood of the observed log viral load from non-failed test results.
The second term on the bottom line is the likelihood of the negative (censored) tests from non-failed test results.

@@

Biological plausibility requires constraining $\vec{\phi}_i$ to positive real values.
Each $\vec{\phi}_i$ is modelled as a log-normal random effects by using the hierarchical prior $\vec{\phi}_i \dist \LN(\vec{\mu}_{\phi,g(i)}, \matr{\Sigma}_\phi)$ where $g(i)$ is the group of individual $i$.
$\vec{\mu}_{\phi,g} = (\mu_{v_{\max},g}, \mu_{a,g}, \mu_{b,g})^T$ is a vector of the population means of the log-parameters.
As the random effects have a log-normal, rather than normal, distribution $\exp(\vec{\mu}_{\phi,g})$ is the population median of the parameter value but not the mean.
The covariance matrix $\matr{\Sigma}_\phi$ controls the variability of the population and allows for the parameters to be correlated.
For example, an individual with a high peak viral load may also clear the virus slowly because both have an underlying cause of a weak immune system.
To improve sampling efficiency, the multivariate version of the non-centred parameterization~\autocite{papaspiliopoulosGeneral,stanReparameterization} is used.
This can be expressed as follows.
\begin{align}
  \vec{\eta}_i &\dist \N(\vec{0}, \matr{C}_\phi) \\
  \log \vec{\phi}_i &= \vec{\mu}_\phi + \vec{\sigma}_\phi \circ \vec{\eta}_i
\end{align}
where $\vec{\sigma}_\phi$ is a vector of the standard deviations of the parameters, and $\matr{C}_\phi$ is a correlation matrix.
These are defined such that, by construction, $(\matr{\Sigma}_\phi)_{ij} = (\vec{\sigma}_\phi)_i (\vec{\sigma}_\phi)_j (\matr{C}_\phi)_{ij}$.
% Correlation matrices have diagonal elements all 1, therefore this reparameterization has the same number of parameters as the original model.

@@

Based on the fit to the data achieved when implementing an earlier version of the model (see \cref{ATACCC:sec:original-analysis}), I hypothesised that the cause of this was overcomplicating the model for test failures.
Specifically, excluding false positives from the model still explains the data well (the results in \cref{fig:paper:duration,ATACCC:fig:goodness-of-fits} do not allow false negatives but fit the data well).
Furthermore, a prior that false negatives and false positives are equally likely is not supported by other information (see \cref{biology-data:sec:measurement-error}).
Therefore, I simplified the model for test failures to only false negatives.
% These false negatives could represent a variety of scenarios, such as poor swabbing technique.
Formally, I replaced \cref{ATACCC:eq:hakki-likelihood} with the following.
\begin{align}
p(y_{i,t} \mid \cdot) &= \begin{cases}
    (1 - \rho) f_N(y_{i,t} \mid \log v_i(t), \sigma_v^2) & \text{if $y_{i,t} > v_\text{lod}$}\\
    \rho + (1 - \rho) F_N(\vlod \mid \log v_i(t), \sigma_v^2) & \text{if $y_{i,t} = v_\text{lod}$.}
\end{cases}
\end{align}

@@

The modified model produced good fits to the data (see \cref{ATACCC:fig:goodness-of-fits}).
Almost all test results are captured in the 95\% credible intervals.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{ATACCC/fits}}
  \caption[Posterior predictive viral load fits]{Posterior predictive observed viral loads (median and 95\% CrI, lines and ribbon) for selected individuals, overlaid with their observed test results (dots). Posterior estimates are conditional on the result not being a false negative. Negative RT-PCR tests are plotted at the limit of detection, \ie a log viral load of 0. Each plot is labelled with the individual's position in the dataset, an arbitrary value. \label{ATACCC:fig:goodness-of-fits}}
\end{figure}

>>>>

First, I will explain how biomarkers are typically modelled, both for SARS-CoV-2 and historically in the biostatistical literature (\cref{ATACCC:sec:background}).
Next, I describe the model of \textcite{hakkiOnset} in detail, which applies these methods to the ATACCC data (\cref{ATACCC:sec:hakki}), before modifying its observation model to improve the robustness of the inference (\cref{ATACCC:sec:observation-modification}).
This modification is based on an earlier analysis using a preliminary version of the dataset, described in \cref{ATACCC:sec:original-analysis}.
I convert the model and parameter estimates of the biomarker's trajectory to a duration of RT-PCR positivity (\cref{ATACCC:sec:duration}).
Finally, I discuss the results (\cref{ATACCC:sec:discussion}).

@@

Random effects represent the within-individual correlation while borrowing information between individuals~\autocite[chapter 24]{lashModern}.
Measurements of viral load from the same individual are correlated.
A random effect means that these measurements share parameter(s), providing correlation.
However, the random effect(s) for each individual is drawn towards a common mean, allowing information on their value to be shared between individuals.
Bayesians often refer to random effect models as hierarchical or multilevel models because they can be interpreted as placing a prior at two levels of a parameter hierarchy~\autocite[chapter 5]{gelmanBDA}.
% $\phi_j$ would be referred to as the bottom level of the hierarchy, with $\mu$ and $\sigma$ being the top level.
% More complex models could have more levels, for instance if the individuals are grouped into subpopulations.

@@

Several features of the \textcite{hakkiOnset} model can be motivated by the desire to apply Stan, a modern and convenient implementation of the NUTS algorithm (see \cref{E-MCMC}).
The first of these is using a continuous viral load trajectory rather than the more common piecewise linear model.
As NUTS uses gradient information, it is more efficient when the model is differentiable.
This leads to the following model for the viral load (see \cref{ATACCC:fig:viral-load-model}):
\begin{align}
v_i(t) = v_{\max,i} \frac{a_i+b_i}{b_ie^{-a_i(t-\tau_{\max,i})} + a_ie^{b_i(t-\tau_{max,i})}}. \label{ATACCC:eq:viral-load}
\end{align}
Time, $t$, is continuous and defined for individual $i$ such that $i$s first test is at $t = 0$.
$v_{\max,i}$ is the maximum viral load, which occurs at time $\tau_{\max,i}$.
As $\lvert t - \tau_{\max} \rvert$ becomes large, the viral load trajectory tends towards exponential growth (before the peak) or decline (after the peak) in viral load at rates $a_i$ and $b_i$ respectively.
% Before the peak, this is because $b(t-\tau_{\max,i})$ is small and hence $a\exp(b(t-\tau_{\max,i}))$ is negligible.
% After the peak, this is because $b\exp(-a(t-\tau_{\max,i}))$ is negligible.
This happens quickly for realistic values of the parameters.
Therefore, the log viral load is approximately piecewise linear (see \cref{ATACCC:fig:viral-load-model} for representative viral load trajectories).
$\vec{\phi}_i = (v_{\max,i}, a_i, b_i)^T$ are the random effects for individual $i$.

@@

The noise in the RT-PCR test result is modelled by assuming each measurement of log viral load has an iid error distributed $\N(0, \sigma_v^2)$.
Measurements are left censored at the limit of detection, $\vlod$, meaning that any observation which would give a log viral load of less than $\vlod$ is instead observed as negative.
False positives and negatives are modelled by allowing a test failure with probability $\rho$.
Test failures are modelled as producing a log viral load reading drawn from $\N(x_0, \sigma_0^2)$.
These readings are censored in the same way as true readings.
A censored test failure represents a false negative, otherwise the test failure represents a false positive.
\textcite{hakkiOnset} uses a prior for $x_0$ centred on $\vlod$, which encodes a prior belief that false negatives and false positives occur with equal probability.

Let the observed log viral load for individual $i$ at time $t$ be $y_{i,t}$, with negative tests signified by $y_{i,t}=v_\text{lod}$.
Then, the likelihood contribution of this observation is:
\begin{align}
p(y_{i,t} \mid \cdot) &= \begin{cases}
    \rho f_N(y_{i,t} \mid x_0, \sigma_0^2) + (1 - \rho) f_N(y_{i,t} \mid \log v_i(t), \sigma_v^2) & \text{if $y_{i,t} > v_\text{lod}$}\\
    \rho F_N(v_\text{lod} \mid x_0, \sigma_0^2) + (1 - \rho) F_N(\vlod \mid \log v_i(t), \sigma_v^2) & \text{if $y_{i,t} = v_\text{lod}$} \label{ATACCC:eq:hakki-likelihood}
\end{cases}
\end{align}
where $f_N(y \mid \mu, \sigma^2)$ and $F_N(y \mid \mu, \sigma^2)$ are the pdf and cdf at $y$ of a normal distribution with mean $\mu$ and variance $\sigma^2$ respectively.
The first term on each line is the likelihood contribution from test failures, either a false positive (top line) or false negative (bottom line).
The second term is the likelihood contribution from non-failed test results, either measurements of viral load (top line) or negative results (bottom line).

@@

Biological plausibility requires constraining $\vec{\phi}_i$ to positive real values.
Each $\vec{\phi}_i$ is modelled as a log-normal random effects by using the hierarchical prior $\vec{\phi}_i \dist \LN(\vec{\mu}_{\phi,g(i)}, \matr{\Sigma}_\phi)$ where $g(i)$ is the group of individual $i$.
$\vec{\mu}_{\phi,g} = [\mu_{v_{\max},g}, \mu_{a,g}, \mu_{b,g}]^T$ is a vector of the population means of the log-parameters.
As the random effects have a log-normal, rather than normal, distribution $\exp(\vec{\mu}_{\phi,g})$ is the population median of the parameter value but not the mean.
The covariance matrix $\matr{\Sigma}_\phi$ controls the variability of the population and allows for the parameters to be correlated.
For example, an individual with a high peak viral load may also clear the virus slowly because both have an underlying cause of a weak immune system.
To improve sampling efficiency, the multivariate version of the non-centred parameterization~\autocite{papaspiliopoulosGeneral,stanReparameterization} is used.
This can be expressed as follows.
\begin{align}
  \vec{\eta}_i &\dist \N(\vec{0}, \matr{C}_\phi) \\
  \log \vec{\phi}_i &= \vec{\mu}_\phi + \vec{\sigma}_\phi \circ \vec{\eta}_i
\end{align}
where $\vec{\sigma}_\phi$ is a vector of the standard deviations of the parameters, and $\matr{C}_\phi$ is a correlation matrix.
These are defined such that, by construction, $(\matr{\Sigma}_\phi)_{ij} = (\vec{\sigma}_\phi)_i (\vec{\sigma}_\phi)_j (\matr{C}_\phi)_{ij}$.
% Correlation matrices have diagonal elements all 1, therefore this reparameterization has the same number of parameters as the original model.

@@

Based on the fit to the data achieved when implementing an earlier version of the model (see \cref{ATACCC:sec:original-analysis}), I hypothesised that the cause of this was overcomplicating the model for test failures.
Specifically, excluding false positives from the model still explains the data well (see \cref{ATACCC:fig:goodness-of-fits}).
Furthermore, a prior that false negatives and false positives are equally likely is not supported by other information (see \cref{biology-data:sec:measurement-error}).
Therefore, I simplified the model for test failures to only false negatives.
% These false negatives could represent a variety of scenarios, such as poor swabbing technique.
Formally, I replaced \cref{ATACCC:eq:hakki-likelihood} with the following.
\begin{align}
p(y_{i,t} \mid \cdot) &= \begin{cases}
    (1 - \rho) f_N(y_{i,t} \mid \log v_i(t), \sigma_v^2) & \text{if $y_{i,t} > v_\text{lod}$}\\
    \rho + (1 - \rho) F_N(\vlod \mid \log v_i(t), \sigma_v^2) & \text{if $y_{i,t} = v_\text{lod}$.}
\end{cases}
\end{align}

@@

The modified model produced good fits to the data (see \cref{ATACCC:fig:goodness-of-fits,ATACCC:fig:appendix-goodness-of-fits}).
Almost all test results are captured in the 95\% credible intervals.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{ATACCC/fits}}
  \caption[Posterior predictive viral load fits]{Posterior predictive observed viral loads (median and 95\% CrI, lines and ribbon) for selected individuals, overlaid with their observed test results (dots). Posterior estimates are conditional on the result not being a false negative. Negative RT-PCR tests are plotted at the limit of detection, \ie a log viral load of 0. Each plot is labelled with the individual's position in the dataset, an arbitrary value.
  Remaining individuals shown in \cref{ATACCC:fig:appendix-goodness-of-fits}.
  \label{ATACCC:fig:goodness-of-fits}}
\end{figure}

### 2024-03-31 MCMC-appendix.tex
proof

The stability of estimates considers the error arising from only using a finite sample of the distribution and the samples being autocorrelated.
Autocorrelation means that estimates using the sample have higher variance than if it were an iid sample from the posterior~\autocite[286]{gelmanBDA}.
The \emph{effective sample size} corrects for the autocorrelation.
The effective sample size can be interpreted as the number of iid samples that would have the same variance as the MCMC sample~\autocites[286]{gelmanBDA}{vehtariRhat}.
Various techniques have been proposed for computing the effective sample size, I will follow the approach of \textcite{vehtariRhat}, implemented in RStan~\autocite{RStan-2-32-3}.

@@

The \emph{No U-Turn Sampler} (NUTS)~\autocite{hoffmanNUTS} is a variant of HMC that automatically tunes the Hamiltonian dynamics hyperparameters.
NUTS has been implemented in a variety of software packages making it easy to apply to arbitrary posterior distributions.
I will make use of the implementation in the package \emph{Sampling through adaptive neighborhoods} (Stan)~\autocite{Stan-2-32-2}.
Stan implements \emph{automatic differentiation}~\autocite{griewankAutoDiff} to calculate the gradient of the posterior density without any user input.
Automatic differentiation uses the chain rule to calculate the gradient of a function from its code by breaking it down into a series of elementary operations, for which the gradient is known~\autocite{autodiff}.
For a more detailed introduction to HMC, NUTS, and Stan see \textcite[chapter 12]{gelmanBDA}.

>>>>

The stability of estimates considers the error arising from only using a finite sample of the distribution and the samples being autocorrelated.
Autocorrelation means that estimates using the sample have higher variance than if it were an iid sample from the posterior~\autocite[286]{gelmanBDA}.
The \emph{effective sample size} corrects for the autocorrelation.
The effective sample size can be interpreted as the number of iid samples that would have the same variance as the MCMC sample~\autocites[286]{gelmanBDA}{vehtariRhat}.
Various techniques have been proposed for computing the effective sample size, I follow the approach of \textcite{vehtariRhat}, implemented in RStan~\autocite{RStan-2-32-3}.

@@

The \emph{No U-Turn Sampler} (NUTS)~\autocite{hoffmanNUTS} is a variant of HMC that automatically tunes the Hamiltonian dynamics hyperparameters.
NUTS has been implemented in a variety of software packages making it easy to apply to arbitrary posterior distributions.
I make use of the implementation in the package \emph{Sampling through adaptive neighborhoods} (Stan)~\autocite{Stan-2-32-2}.
Stan implements \emph{automatic differentiation}~\autocite{griewankAutoDiff} to calculate the gradient of the posterior density without any user input.
Automatic differentiation uses the chain rule to calculate the gradient of a function from its code by breaking it down into a series of elementary operations, for which the gradient is known~\autocite{autodiff}.
For a more detailed introduction to HMC, NUTS, and Stan see \textcite[chapter 12]{gelmanBDA}.

### 2024-03-31 biology-data.tex
proof

SARS-CoV-2 is closely related to previous coronaviruses that have caused outbreaks in humans, including SARS-CoV and MERS-CoV.
Coronaviruses are RNA (Ribonucleic acid) viruses which means they use RNA, rather than DNA (Deoxyribonucleic acid), as their genetic material.
RNA viruses, which include influenza, mutate rapidly and are the most likely class of viruses to cause pandemics~\autocite{woolhouseRNA}.

@@

SARS-CoV-2 is an RNA virus, but PCR amplifies DNA.
Therefore, before using PCR process to detect SARS-CoV-2, the RNA must be converted to DNA.
This process is known as \emph{reverse transcription}.
In reverse transcription, a naturally occurring enzyme known as reverse transcriptase generates cDNA (complementary DNA) from the RNA in the sample~\autocite{valasekPower}.
Any sequence of cDNA has a one-to-one correspondence with a sequence of RNA.
The cDNA can then be used in the standard PCR process.

@@

A host's \emph{viral load}, the quantity of viral particles in a host, is the underlying biological process driving the different periods and characteristics of an infection~\autocite{puhachSARSCoV2}.
Any host's viral load is determined by the interaction of many biological processes, most prominently the virus's replication increasing the viral load and the host's immune response reducing the viral load~\autocite{keVivo}.
Viral replication requires the virus to infect a host cell and then use the host cell's machinery to produce more virus; as more host cells are infected this process slows because there are fewer remaining potential host cells, known as \emph{target cells}, to infect.
The host's immune system can modify this process by various mechanisms; the most relevant for SARS-CoV-2 is by making some target cells temporarily immune from infection.
The immune system response contribution to the reduction in target cells is several orders of magnitude larger than the reduction in target cells due to viral replication.
The rate at which host cells are infected slows over time because there are fewer remaining target cells to infect.
A host's immune response is triggered by the viral load and becomes stronger as the viral load increases.
A high viral load is associated with being more infectious, more likely to test positive on an RT-PCR test, and the start of symptoms~\autocite{puhachSARSCoV2,keVivo}.

@@

The ATACCC study was a longitudinal study of contacts reported to the NHS Test and Trace contact tracing service~\autocite{hakkiOnset}.
Contacts were invited to participate if they were 5 years or older and  could be reached within 5 days of the index case's symptom onset date.
ATACCC enrolled a total of 738 contacts across two waves.
The first wave enrolled 393 contacts between 13 Sep 2020 and 31 Mar 2021, when pre-Alpha and Alpha variants were circulating.
The second wave enrolled 345 contacts between 24 May 2021 and 28 Oct 2021, when the Delta variant was circulating.
Contacts were swabbed daily for up to 20 days, with a variety of tests performed on the swabs; relevant for this thesis is the RT-PCR test for SARS-CoV-2.
For all positive test results, the viral load was measured using the Ct value and then converted to a viral load using a standard curve (see \cref{biology-data:sec:PCR-process}).
Other data such as demography and symptoms were also collected, although I do not use this information.
For further details see \textcite{singanayagamDuration,hakkiOnset}.

@@

\begin{figure}
  \centering \includegraphics{biology-data/CIS-recruitment}
  \caption[CIS swab numbers]{%
    Number of swabs taken by the CIS in the period considered in this thesis (Sep 2020 until Jan 2021).
    (A) Only enrolment (\ie first) swabs for each individual.
    (B) Total swabs.
    Note the differing y-axis scales.
    Data from \textcite{CIStechData}.
  }
  \label{biology-data:fig:CIS-recruitment}
\end{figure}
% \begin{figure}
%   \centering \includegraphics[width=\textwidth]{biology-data/CIS-num-tests}
%   \caption[Number of CIS tests]{%
%     Daily number of CIS tests conducted in the period considered in this thesis (Sep 2020 until Jan 2021).
%   }
%   \label{biology-data:fig:CIS-num-tests}
%   \todo[inline]{Cut this repetition of number of tests? Maybe highlight period of interest in previous plot}
% \end{figure}

@@

I use two levels of ethnic grouping.
Broad groupings (White and non-White), because the White population makes up a large majority of the UK.
I also use more granular ethnicities (Asian, Black, Mixed, White, and Other).

@@

If the CIS was truly a random sample, an unbiased estimator of the population positivity rate would be the proportion of positive tests in the sample; I refer to this estimator as the unadjusted positivity rate.
However, in practice, two issues arise with this.
First, the non-response (discussed above) causes bias.
Second, the number of positives on any day is low, especially among any specific subpopulation (\eg age group or region) increases variance.
Statistical methods are used to correct for these issues.

@@

The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
To start, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution.
Denote these by $p^{(j)}_{t,i}$ for the $j$th sample where $i$ is a stratum, \ie some value of $a$, $e$, and $s$.
Define a (sub) population of interest as a set of strata $\set{S}$.
Then the prevalence in the (sub)population on day $t$ is
\begin{align}
p^{(j)}_t(\set{S})
&= \frac{\sum_{i \in \set{S}} p^{(j)}_{t,i} N_i}{\sum_{i \in \set{S}} N_i}
\end{align}
where $N_i$ is the population size of stratum $i$.
The population numbers for each stratum are provided by the ONS.

@@

\Cref{biology-data:fig:CIS-positivity} shows estimates from this model compared to the unadjusted positivity rate.
Both show similar trends.
The MRP estimate is normally greater than the uncorrected positivity rate because the underrepresented demographic groups tended to have higher positivity rates.

>>>>

SARS-CoV-2 is closely related to previous coronaviruses that have caused outbreaks in humans, including SARS-CoV and MERS-CoV.
Coronaviruses are RNA (Ribonucleic acid) viruses, which means they use RNA, rather than DNA (Deoxyribonucleic acid), as their genetic material.
RNA viruses, which include influenza, mutate rapidly and are the most likely class of viruses to cause pandemics~\autocite{woolhouseRNA}.

@@

SARS-CoV-2 is an RNA virus, but PCR amplifies DNA.
Therefore, before using the PCR process to detect SARS-CoV-2, the RNA must be converted to DNA.
This process is known as \emph{reverse transcription}.
In reverse transcription, a naturally occurring enzyme known as \emph{reverse transcriptase} generates cDNA (complementary DNA) from the RNA in the sample~\autocite{valasekPower}.
Any sequence of cDNA has a one-to-one correspondence with a sequence of RNA.
The cDNA can then be used in the standard PCR process.

@@

A host's \emph{viral load}, the quantity of viral particles in a host, is the underlying biological process driving the different periods and characteristics of an infection~\autocite{puhachSARSCoV2}.
Any host's viral load is determined by the interaction of many biological processes, most prominently the virus's replication increasing the viral load and the host's immune response reducing the viral load~\autocite{keVivo}.
Viral replication requires the virus to infect a host cell and then use the host cell's machinery to produce more virus; as more host cells are infected this process slows because there are fewer remaining potential host cells, known as \emph{target cells}, to infect.
The host's immune system can modify this process by various mechanisms; the most relevant for SARS-CoV-2 is by making some target cells temporarily immune from infection.
The immune system response contribution to the reduction in target cells is several orders of magnitude larger than the reduction in target cells due to viral replication.
The response is triggered by the viral load and becomes stronger as the viral load increases.
The rate at which host cells are infected slows over time because there are fewer remaining target cells to infect.
A high viral load is associated with being more infectious, more likely to test positive on an RT-PCR test, and the start of symptoms~\autocite{puhachSARSCoV2,keVivo}.

@@

The ATACCC study was a longitudinal study of contacts reported to the NHS Test and Trace contact tracing service~\autocite{hakkiOnset}.
Contacts were invited to participate if they were 5 years or older and  could be reached within 5 days of the index case's symptom onset date.
ATACCC enrolled a total of 738 contacts across two waves.
The first wave enrolled 393 contacts between 13 Sep 2020 and 31 Mar 2021, when pre-Alpha and Alpha variants were circulating.
The second wave enrolled 345 contacts between 24 May 2021 and 28 Oct 2021, when the Delta variant was circulating.
Contacts were swabbed daily for up to 20 days, with a variety of tests performed on the swabs; relevant for this thesis is the RT-PCR test for SARS-CoV-2.
For all positive test results, the viral load was measured using the Ct value and then converted to a viral load using a standard curve (see \cref{biology-data:sec:PCR-process}).
Other data such as demography and symptoms were also collected.
For further details see \textcite{singanayagamDuration,hakkiOnset}.

@@

\begin{figure}
  \centering \includegraphics{biology-data/CIS-recruitment}
  \caption[CIS swab numbers]{%
    Number of swabs taken by the CIS up until Jan 2021.
    (A) Only enrolment (\ie first) swabs for each individual.
    (B) Total swabs.
    Note the differing y-axis scales.
    Data from \textcite{CIStechData}.
  }
  \label{biology-data:fig:CIS-recruitment}
\end{figure}
% \begin{figure}
%   \centering \includegraphics[width=\textwidth]{biology-data/CIS-num-tests}
%   \caption[Number of CIS tests]{%
%     Daily number of CIS tests conducted in the period considered in this thesis (Sep 2020 until Jan 2021).
%   }
%   \label{biology-data:fig:CIS-num-tests}
%   \todo[inline]{Cut this repetition of number of tests? Maybe highlight period of interest in previous plot}
% \end{figure}

@@

I use two levels of ethnic grouping.
Broad groupings (White and non-White), because the White population makes up over 80\% of the UK.
I also use more granular ethnicities (Asian, Black, Mixed, White, and Other).

@@

If the CIS was truly a sample representative of the general population, an unbiased estimator of the population positivity rate would be the proportion of positive tests in the sample; I refer to this estimator as the unadjusted positivity rate.
However, in practice, two issues arise with this.
First, the non-response (discussed above) causes bias.
Second, the number of positives on any day is low, especially among any specific subpopulation (\eg age group or region) increases variance.
Statistical methods are used to correct for these issues.

@@

The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
To start, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution.
Denote these by $p^{(j)}_{t,i}$ for the $j$th sample where $i$ is a stratum, \ie some value of $a$, $e$, and $s$.
Define a (sub) population of interest as a set of strata $\set{S}$.
Then the poststratified prevalence in the (sub)population on day $t$ is
\begin{align}
p^{(j)}_t(\set{S})
&= \frac{\sum_{i \in \set{S}} p^{(j)}_{t,i} N_i}{\sum_{i \in \set{S}} N_i}
\end{align}
where $N_i$ is the population size of stratum $i$.
The population numbers for each stratum are provided by the ONS.

@@

\Cref{biology-data:fig:CIS-positivity} shows estimates from this model compared to the unadjusted positivity rate.
Both show similar trends.
The MRP estimate is normally greater than the unadjusted positivity rate because the underrepresented demographic groups tended to have higher positivity rates.
